---
title: "Predicting EMS call response time in Virginia Baech"
author: "David Seunglee Park"
date: "12/10/2020"
output: 
  html_document:
    code_folding: hide
    theme: cosmo
    highlight: tango
    toc: true
    toc_float:
      collapsed: false


---

# 1. Introduction

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load libraries
library(tidyverse)
library(tidycensus)
library(dplyr)
library(sf)
library(mapview)
library(viridis) 
library(riem)
library(lubridate)
library(gridExtra)
library(leaflet)
library(shiny)
library(mapboxapi)
library(riem)
library(measurements)
library(mapboxapi)
library(FNN)
library(kableExtra)
```

```{r theme and functions, echo=FALSE}
#themes and nn_function
# Set up plot and map themes
plotTheme <- theme(text = element_text( family = "Avenir", color = "black"),
                   plot.title =element_text(size=13, face = "bold"),
                   plot.subtitle = element_text(size=8),
                   plot.caption = element_text(size = 6),
                   axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
                   axis.title.x = element_text(size = 10),
                   axis.text.y = element_text(size = 8),
                   axis.title.y = element_text(size = 10),
                   # Set the entire chart region to blank
                   panel.background=element_blank(),
                   plot.background=element_blank(),
                   #panel.border=element_rect(colour="#F0F0F0"),
                   # Format the grid
                   panel.grid.major=element_line(colour="#565050",size=.2),
                   axis.ticks=element_blank())

mapTheme <- theme(text = element_text( family = "Avenir", color = "black"),
                  plot.title =element_text(size=20),
                  plot.subtitle = element_text(size=7),
                  plot.caption = element_text(size = 5),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]]), digits = 3),
                 c(.01,.2,.4,.6,.8), na.rm=T)
  }
}


q5 <- function(variable) {as.factor(ntile(variable, 5))}

nn_function <- function(measureFrom,measureTo,k) {
  measureFrom_Matrix <-
    as.matrix(measureFrom)
  measureTo_Matrix <-
    as.matrix(measureTo)
  nn <-   
    get.knnx(measureTo_Matrix, measureFrom_Matrix, k)$nn.dist
  output <-
    as.data.frame(nn) %>%
    rownames_to_column(var = "thisPoint") %>%
    gather(points, point_distance, V1:ncol(.)) %>%
    arrange(as.numeric(thisPoint)) %>%
    group_by(thisPoint) %>%
    summarize(pointDistance = mean(point_distance)) %>%
    arrange(as.numeric(thisPoint)) %>% 
    dplyr::select(-thisPoint) %>%
    pull()
  
  return(output)  
}

# kloat color palettes
palette5 <- c("#03254c","#1167b1","#187bcd","#2a9df4","#d0efff")
palette4 <- c("#3a7d7c","#ffa137","#ff4400","#065125")
palette2 <- c("#03254c","#187bcd")
palette1 <- c("#03254c")

my_token <- "pk.eyJ1IjoiZ2F1bHQzNCIsImEiOiJja2ZsbWd5cm8xNDBsMnlwajMzbW15c2Y0In0.nZ9siGKFAjMx_JQVEzeOtg"
```


# 2. Data

```{r}
datasource_table <-data.frame(
  Datasets = c("Virginia Beach EMS calls data", "Tract level U.S. Census demographic data", "Weather data", "Tract level chronic disease and health outcome data"),
  Source = c("Provided by class", "U.S. Census Bureau", "Riem package which gather weather data from \n Automated Surface Observing System (ASOS) stations (airports)", "Center for Disease Control and Prevention")
)

kbl(datasource_table, caption ="<b>Datasets used<b>") %>%
  kable_paper(c("hover", "condensed"), full_width = F, html_font = "Avenir") %>%
  kable_styling (bootstrap_options = "striped", "condensed", font_size = 10) %>%
  row_spec(0, bold = T, color = "white", background = "dodgerblue", font_size = 14) %>%
  column_spec(1, bold = T)
```

### 2.1. Loading main EMS data
```{r, load the main dataset}
# load main dataset and make it an SF object
main_ems <- read.csv("Main_VaBeach_EMS_2017_18.csv")
main_ems.sf <- main_ems %>%
  na.omit() %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
  st_transform('EPSG:6595')
```

```{r, time feature table, results='asis'}
timefeatures_table <-data.frame(
  Features = c("ResponseTime", "time_of_day", "times", "dotw", "weekend", "CallVolume"),
  Description = c("Time from the initial call to EMS arriving at scene", "6 hour periods (Overnight, Morning, Afternoon, Evening)", "3 hour periods (one-three, three-six, six-nine, nine-twelve, twelve-fifteen, fifteen-eighteen, eighteen-twentyone,twentyone-twentyfour)", "Day of the week", "weekend", "Number of calls during the 3 hour period"),
  Type = c("Continous","Categorical", "Categorical", "Categorical", "Binary", "Continous")
)

kbl(timefeatures_table, caption ="<b>Engineered temporal features<b>") %>%
  kable_paper(c("hover", "condensed"), full_width = F, html_font = "Avenir") %>%
  kable_styling (bootstrap_options = "striped", "condensed", font_size = 10) %>%
  row_spec(0, bold = T, color = "white", background = "dodgerblue", font_size = 14) %>%
  column_spec(1, bold = T)
```

```{r, feature engineer time vars}
# create time bins (30 min and 60 min)
# create days and weeks
# create ResponseTime from "CallDateandTime" to "OnSceneDateandTime" 
# create "time_of_day" (6 hour periods)
# create "times" (3 hour periods)
main_ems.sf <- main_ems.sf %>%
  mutate(interval60 = floor_date(mdy_hm(CallDateandTime), unit = "60 mins"),
         interval30 = floor_date(mdy_hm(CallDateandTime), unit = "30 mins"),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>%
  mutate(ResponseTime =  as.numeric(difftime(mdy_hm(OnSceneDateandTime), mdy_hm(CallDateandTime), units = "min"))) %>%
  mutate(time_of_day = case_when(hour(interval60) >= 0 & hour(interval60) < 6 ~ "Overnight",
                                 hour(interval60) >= 6 & hour(interval60) < 12 ~ "Morning",
                                 hour(interval60) >= 12 & hour(interval60) < 18 ~ "Afternoon",
                                 hour(interval60) >= 18 & hour(interval60) <= 24 ~ "Evening")) %>%
  mutate(times = case_when(hour(interval60) >= 0 & hour(interval60) < 3 ~ "one-three",
                           hour(interval60) >= 3 & hour(interval60) < 6 ~ "three-six",
                           hour(interval60) >= 6 & hour(interval60) < 9 ~ "six-nine",
                           hour(interval60) >= 9 & hour(interval60) < 12 ~ "nine-twelve",
                           hour(interval60) >= 12 & hour(interval60) < 15 ~ "twelve-fifteen",
                           hour(interval60) >= 15 & hour(interval60) < 18 ~ "fifteen-eighteen",
                           hour(interval60) >= 18 & hour(interval60) < 21 ~ "eighteen-twentyone",
                           hour(interval60) >= 21 & hour(interval60) <= 24 ~ "twentyone-twentyfour")) %>%
  mutate(weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday")) %>%
  mutate(Date = substring(CallDateandTime,1,6)) %>%
  unite(Date_timeofday, c(Date, time_of_day), sep = " ", remove = FALSE) %>%
  unite(Date_time, c(Date, times), sep = " ", remove = FALSE)

```

```{r, call volume}
# create call volume column
vol_count_dat <- main_ems.sf %>%
  st_drop_geometry() %>%
  group_by(Date, times) %>%
  summarise(CallVolume = n()) %>%
  unite(Date_time, c(Date, times), sep = " ", remove = FALSE)

# join the volume data file to main dataframe
main_ems.sf <-
  left_join(main_ems.sf, vol_count_dat, by="Date_time")
```

Rescue Squad type feature engineering
```{r, rescue squad feature table}
RescueSquad_table <-data.frame(
  Features = c("advanced_life_support", "chief_dispatch", "special_event", "air_dispatch", "hold_dispatch", "water_dispatch", "mass_casualty_dispatch", "fire_dispatch"),
  Description = c("Calls requiring advanced life support unit", "Calls requiring an EMS chief","Calls at a special event, like a concert or marathon", "Calls requiring Helicopter Ambulances", "Calls when the dispatcher was waiting on another ambulence", "Marine EMS Calls requiring rescue boat or jetski", "Calls requiring mass causalty unit", "Calls related to fire incidents"),
  Type = c("Binary","Binary", "Binary", "Binary", "Binary", "Binary", "Binary", "Binary")
)

kbl(RescueSquad_table , caption ="<b>Rescue squad types engineered features<b>") %>%
  kable_paper(c("hover", "condensed"), full_width = F, html_font = "Avenir") %>%
  kable_styling (bootstrap_options = "striped", "condensed", font_size = 10) %>%
  row_spec(0, bold = T, color = "white", background = "dodgerblue", font_size = 14) %>%
  column_spec(1, bold = T)
```

```{r, rescue squad type}
main_ems.sf$advanced_life_support <- ifelse (
  (
    endsWith(main_ems.sf$RescueSquad.Number, 'S') |
      startsWith(main_ems.sf$RescueSquad.Number, 'Z') |
      endsWith(main_ems.sf$RescueSquad.Number, 'P') |
      startsWith(main_ems.sf$RescueSquad.Number, 'MED')
  ), 
  "Advanced Life Support", "Not Advanced Life Support"
)

main_ems.sf$b_advanced_life_support <- ifelse (
  (
    endsWith(main_ems.sf$RescueSquad.Number, 'S') |
      startsWith(main_ems.sf$RescueSquad.Number, 'Z') |
      endsWith(main_ems.sf$RescueSquad.Number, 'P') |
      startsWith(main_ems.sf$RescueSquad.Number, 'MED')
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(advanced_life_support) %>%
         drop_na(advanced_life_support) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=advanced_life_support, y=meanResponeTime, fill=advanced_life_support), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of EMS Calls requiring Advanced Life Support") +
  plotTheme

main_ems.sf$chief_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'ECH') |
      (((startsWith(main_ems.sf$RescueSquad.Number, 'CAR')) == TRUE) & ((startsWith(main_ems.sf$RescueSquad.Number, 'CART')) == FALSE)) |
      startsWith(main_ems.sf$RescueSquad.Number, 'BAT')
  ), 
  "Chief Dispatched", "Cheif not Dispatched"
)

main_ems.sf$b_chief_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'ECH') |
      (((startsWith(main_ems.sf$RescueSquad.Number, 'CAR')) == TRUE) & ((startsWith(main_ems.sf$RescueSquad.Number, 'CART')) == FALSE)) |
      startsWith(main_ems.sf$RescueSquad.Number, 'BAT')
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(chief_dispatch) %>%
         drop_na(chief_dispatch) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=chief_dispatch, y=meanResponeTime, fill=chief_dispatch), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of EMS Calls needing a Chief") +
  plotTheme

main_ems.sf$special_event <- ifelse (
  (
    main_ems.sf$RescueSquad.Number == 'BKTEAM' |
      main_ems.sf$RescueSquad.Number == 'EMSOPS' |
      startsWith(main_ems.sf$RescueSquad.Number, 'CART')
  ), 
  "Special Event", "Not Special Event"
)

main_ems.sf$b_special_event <- ifelse (
  (
    main_ems.sf$RescueSquad.Number == 'BKTEAM' |
      main_ems.sf$RescueSquad.Number == 'EMSOPS' |
      startsWith(main_ems.sf$RescueSquad.Number, 'CART')
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(special_event) %>%
         drop_na(special_event) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=special_event, y=meanResponeTime, fill=special_event), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of EMS Calls at Special Events") +
  plotTheme

main_ems.sf$air_dispatch <- ifelse (
  (
    main_ems.sf$RescueSquad.Number == 'AIRMED'
  ), 
  "Air unit", "Not an Air Unit"
)

main_ems.sf$b_air_dispatch <- ifelse (
  (
    main_ems.sf$RescueSquad.Number == 'AIRMED'
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(air_dispatch) %>%
         drop_na(air_dispatch) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=air_dispatch, y=meanResponeTime, fill=air_dispatch), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of EMS Calls Using Helicopter Ambulances") +
  plotTheme

main_ems.sf$hold_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'HOLD')
  ), 
  "Held By Dispatcher", "Not Held by Dispatcher"
)

main_ems.sf$b_hold_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'HOLD')
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(hold_dispatch) %>%
         drop_na(hold_dispatch) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=hold_dispatch, y=meanResponeTime, fill=hold_dispatch), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) +
  labs(title= "Average Response Time Difference of EMS Calls on HOLD by Dispatcher") +
  plotTheme

main_ems.sf$water_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'FBOA') |
      startsWith(main_ems.sf$RescueSquad.Number, 'RB') |
      startsWith(main_ems.sf$RescueSquad.Number, 'MRTK') |
      startsWith(main_ems.sf$RescueSquad.Number, 'JTSKI')
    
  ), 
  "Water Unit", "Not a Water Unit"
)

main_ems.sf$b_water_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'FBOA') |
      startsWith(main_ems.sf$RescueSquad.Number, 'RB') |
      startsWith(main_ems.sf$RescueSquad.Number, 'MRTK') |
      startsWith(main_ems.sf$RescueSquad.Number, 'JTSKI')
    
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(water_dispatch) %>%
         drop_na(water_dispatch) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=water_dispatch, y=meanResponeTime, fill=water_dispatch), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of Marine EMS Calls") +
  plotTheme

main_ems.sf$mass_casualty_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'MCI') 
  ), 
  "Mass Casualty Incident", "Not a Mass Casualty Incident"
)

main_ems.sf$b_mass_casualty_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'MCI') 
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(mass_casualty_dispatch) %>%
         drop_na(mass_casualty_dispatch) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=mass_casualty_dispatch, y=meanResponeTime, fill=mass_casualty_dispatch), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of Mass Casualty EMS Calls") +
  plotTheme

main_ems.sf$fire_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'L') |
      (startsWith(main_ems.sf$RescueSquad.Number, 'E') & ((startsWith(main_ems.sf$RescueSquad.Number, 'ECH') == FALSE) & (main_ems.sf$RescueSquad.Number != 'EMSOPS'))) |
      (startsWith(main_ems.sf$RescueSquad.Number, 'T') & (startsWith(main_ems.sf$RescueSquad.Number, 'TAC') == FALSE)) |
      endsWith(main_ems.sf$RescueSquad.Number, 'F') |
      startsWith(main_ems.sf$RescueSquad.Number, 'NE')
  ), 
  "Fire Team Dispactched", "Fire Team not Dispatched"
)

main_ems.sf$b_fire_dispatch <- ifelse (
  (
    startsWith(main_ems.sf$RescueSquad.Number, 'L') |
      (startsWith(main_ems.sf$RescueSquad.Number, 'E') & ((startsWith(main_ems.sf$RescueSquad.Number, 'ECH') == FALSE) & (main_ems.sf$RescueSquad.Number != 'EMSOPS'))) |
      (startsWith(main_ems.sf$RescueSquad.Number, 'T') & (startsWith(main_ems.sf$RescueSquad.Number, 'TAC') == FALSE)) |
      endsWith(main_ems.sf$RescueSquad.Number, 'F') |
      startsWith(main_ems.sf$RescueSquad.Number, 'NE')
  ), 
  1, 0
)

ggplot(data = main_ems.sf %>%
         group_by(fire_dispatch) %>%
         drop_na(fire_dispatch) %>%
         summarise(meanResponeTime = mean(ResponseTime, na.rm=TRUE))) +
  geom_bar(aes(x=fire_dispatch, y=meanResponeTime, fill=fire_dispatch), stat="identity", position=position_dodge(), show.legend = FALSE) +
  scale_fill_manual(values = palette2) + 
  labs(title= "Average Response Time Difference of Fire-related EMS Calls") +
  plotTheme
```

### 2.4 Load spatial features


```{r}
nn_table <-data.frame(
  Features = c("nn_ems_station", "nn_fire_stations", "nn_hospitals"),
  Description = c("distance to nearest EMS station", "distance to nearest fire station", "distance to nearest hosptial"),
  Type = c("Continuous","Continuous", "Continuous")
)

kbl(nn_table , caption ="<bEengineered distance features<b>") %>%
  kable_paper(c("hover", "condensed"), full_width = T, html_font = "Avenir") %>%
  kable_styling (bootstrap_options = "striped", "condensed", font_size = 10) %>%
  row_spec(0, bold = T, color = "white", background = "dodgerblue", font_size = 14) %>%
  column_spec(1, bold = T)
```

```{r, loading map files}
# load VB boundary census tracts
vb_boundary <-
  st_read("https://gismaps.vbgov.com/arcgis/rest/services/Basemaps/Administrative_Boundaries/MapServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json") %>%
  st_transform('EPSG:6595')

# load VB tracts
vb_tracts <- 
  st_read("https://opendata.arcgis.com/datasets/82ada480c5344220b2788154955ce5f0_2.geojson") %>%
  subset(OBJECTID!= 22) %>%
  st_transform('EPSG:6595')

# create fishnet
vb_fishnet <- 
  st_make_grid(vb_boundary,
               cellsize = 1000, 
               square = FALSE) %>%
  .[vb_boundary] %>% 
  st_sf() %>%
  mutate(uniqueID = rownames(.))

#load ems stations hospitals and fire stations
ems_stations <- st_read('201127_ems_geocoded.shp') %>%
  st_transform(st_crs(vb_fishnet)) %>%
  mutate(OBJECTID = osm_id) %>%
  mutate(Legend = "ems_stations")

hospitals <- st_read('Hospitals__Virginia_.shp') %>% 
  dplyr::filter(City == 'Virginia Beach') %>%
  st_transform(st_crs(vb_fishnet)) %>%
  mutate(Legend = "hospitals")

fire_stations <- st_read('Fire_Stations.shp') %>% 
  dplyr::filter(CITY == 'Virginia Beach') %>%
  st_transform(st_crs(vb_fishnet)) %>%
  mutate(Legend = "fire_stations")
```


```{r}
# create voronoi (still needs work)
bbox_polygon <- function(x) {
  bb <- sf::st_bbox(x)
  
  p <- matrix(
    c(bb["xmin"], bb["ymin"], 
      bb["xmin"], bb["ymax"],
      bb["xmax"], bb["ymax"], 
      bb["xmax"], bb["ymin"], 
      bb["xmin"], bb["ymin"]),
    ncol = 2, byrow = T
  )
  
  sf::st_polygon(list(p))
}

box <- st_sfc(bbox_polygon(vb_boundary))

v <- st_voronoi(st_union(ems_stations), box)
plot(v, col = 0)

voronoi_polygon <- (st_intersection(st_cast(v), st_union(vb_boundary)))

plot(voronoi_polygon)

#convert sfc to sf and assign ID
vb_voronoi = voronoi_polygon %>%
  st_sf %>%
  st_cast %>%
  dplyr::mutate(Voronoi_ID = row_number())

#Assign Voronoi ID for each EMS call
main_ems.sf <- st_join(main_ems.sf, vb_voronoi, left = TRUE)
```



### 2.3 Load census data

```{r}
census_vars_table <-data.frame(
  Features = c("Med_Age", "MedHHInc", "PopDens", "pctWhite", "pctBlack", "pctHis", "pctBachelors","pctPoverty", "pctCarCommute", "pctPubCommute"),
  Description = c("Median Age", "Median household income", "Population Density", "Percentage of White population", "Percentage of Black population", "Percentage of Hispanic populaiton", "Percentage of population with at least Bachelor's degree", "Percentage of population below the poverty line", "Percentage of population communiting by car", "Percentage of population commuting by public transit"),
  "Data type" = c("continuous", "continuous", "continuous", "continuous", "continuous", "continuous", "continuous", "continuous", "continuous", "continuous"))

kbl(census_vars_table , caption ="<b> U.S. Census features<b>") %>%
  kable_paper(c("hover", "condensed"), full_width = T, html_font = "Avenir") %>%
  kable_styling (bootstrap_options = "striped", "condensed", font_size = 10) %>%
  row_spec(0, bold = T, color = "white", background = "dodgerblue", font_size = 14) %>%
  column_spec(1, bold = T)
```


```{r,load census data}
# Load census (pop_density, med_age, race, poverty, income, education, commute)
census_api_key("41e1c0d912341017fa6f36a5da061d3b23de335e", overwrite = TRUE)
selected_vars <- c("B01003_001E", # Total Population
                   "B02001_001E", # Estimate!!Total population by race -- ##let's double check that it's okay to use this as long as we justify it
                   "B02001_002E", # People describing themselves as "white alone"
                   "B02001_003E", # People describing themselves as "black" or "african-american" alone
                   "B15001_050E", # Females with bachelors degrees
                   "B15001_009E", # Males with bachelors degrees
                   "B19013_001E", # Median HH income
                   "B25058_001E", # Median rent
                   "B06012_002E", # Total poverty
                   "B08301_001E", # People who have means of transportation to work
                   "B08301_002E", # Total people who commute by car, truck, or van
                   "B08301_010E", # Total people who commute by public transportation"
                   "B03002_012E", # Estimate Total Hispanic or Latino by race
                   "B19326_001E", # Median income in past 12 months (inflation-adjusted)
                   "B07013_001E", # Total households
                   "B08013_001E", # Travel Time to Work
                   "B01002_001E") # Median Age

vb_census <- 
  get_acs(geography = "tract", 
          variables = selected_vars, 
          year=2018, 
          state="VA",
          county = c("Virginia Beach"),
          geometry=T, 
          output="wide") %>%
  st_transform('EPSG:6595') %>%
  rename(TotalPop = B01003_001E,
         Med_Age = B01002_001E,
         Race_TotalPop = B02001_001E, 
         Whites = B02001_002E,
         Blacks = B02001_003E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E,
         TotalPoverty = B06012_002E,
         TotalCommute = B08301_001E,
         CarCommute = B08301_002E,
         PubCommute = B08301_010E,
         TotalHispanic = B03002_012E,
         MedInc = B19326_001E,
         TotalHH = B07013_001E)
vb_census <- vb_census %>%
  dplyr::select(-NAME, -starts_with("B0"), -starts_with("B1"), -starts_with("B2")) %>%
  mutate(Area = st_area(vb_census),
         pctWhite = (ifelse(Race_TotalPop > 0, Whites / Race_TotalPop,0))*100,
         pctBlack = (ifelse(Race_TotalPop > 0, Blacks / Race_TotalPop,0))*100,
         pctHis = (ifelse(Race_TotalPop >0, TotalHispanic/Race_TotalPop,0))*100,
         pctBachelors = (ifelse(Race_TotalPop > 0, ((FemaleBachelors + MaleBachelors) / Race_TotalPop),0)) *100,
         pctPoverty = (ifelse(Race_TotalPop > 0, TotalPoverty / Race_TotalPop, 0))*100,
         pctCarCommute = (ifelse(TotalCommute > 0, CarCommute / TotalCommute,0))*100,
         pctPubCommute = (ifelse(TotalCommute > 0, PubCommute / TotalCommute,0))*100,
         year = "2018") %>%
  mutate(MedHHInc = replace_na(MedHHInc, 0),
         pctBachelors= replace_na(pctBachelors,0),
         pctHis= replace_na(pctHis,0),
         pctCarCommute= replace_na(pctCarCommute,0),
         PopDens = (TotalPop/(Area/27878400))) %>%
  dplyr::select(-Whites, -Blacks, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -CarCommute, -PubCommute, -TotalCommute, -TotalHispanic)
```


```{r, load helth data}
health_dat <- read.csv('health_data_500_cities_vabch.csv') %>% 
  dplyr::select(TractFIPS, ends_with('CrudePrev')) %>% 
  rename(GEOID = TractFIPS)
```

```{r, merging to main daraframe}
vb_health <- merge(health_dat, vb_census,
                   by.x = "GEOID", by.y = "GEOID",
                   all.x = FALSE, all.y = TRUE,
                   sort = FALSE) %>% 
  dplyr::select(GEOID, ends_with('CrudePrev'), geometry) %>% 
  st_sf()


main_ems.sf <-st_join(main_ems.sf,vb_census, left =TRUE) %>%
  mutate(TotalPop = TotalPop,
         MedHHInc = MedHHInc,
         TotalHH = TotalHH,
         pctWhite = pctWhite,
         pctBlack = pctWhite,
         pctHis= pctHis,
         pctBachelors= pctBachelors,
         pctPoverty = pctPoverty,
         pctCarCommute= pctCarCommute,
         pctPubCommute= pctPubCommute,
         PopDens= PopDens)


#you can find variable names here: https://www.cdc.gov/places/about/500-cities-2016-2019/index.html
main_ems.sf <-st_join(main_ems.sf,vb_health, left =TRUE) %>%
  mutate(
    ACCESS2_CrudePrev = ACCESS2_CrudePrev,
    ARTHRITIS_CrudePrev = ARTHRITIS_CrudePrev,
    BINGE_CrudePrev = BINGE_CrudePrev,
    BPHIGH_CrudePrev = BPHIGH_CrudePrev,
    BPMED_CrudePrev = BPMED_CrudePrev,
    CANCER_CrudePrev = CANCER_CrudePrev,
    CASTHMA_CrudePrev = CASTHMA_CrudePrev,
    CHD_CrudePrev = CHD_CrudePrev,
    CHECKUP_CrudePrev = CHECKUP_CrudePrev,
    CHOLSCREEN_CrudePrev = CHOLSCREEN_CrudePrev,
    COLON_SCREEN_CrudePrev = COLON_SCREEN_CrudePrev,
    COPD_CrudePrev = COPD_CrudePrev,
    COREM_CrudePrev = COREM_CrudePrev,
    COREW_CrudePrev = COREW_CrudePrev,
    CSMOKING_CrudePrev = CSMOKING_CrudePrev,
    DENTAL_CrudePrev = DENTAL_CrudePrev,
    DIABETES_CrudePrev = DIABETES_CrudePrev,
    HIGHCHOL_CrudePrev = HIGHCHOL_CrudePrev,
    KIDNEY_CrudePrev = KIDNEY_CrudePrev,
    LPA_CrudePrev = LPA_CrudePrev,
    MAMMOUSE_CrudePrev = MAMMOUSE_CrudePrev,
    MHLTH_CrudePrev = MHLTH_CrudePrev,
    OBESITY_CrudePrev = OBESITY_CrudePrev,
    PAPTEST_CrudePrev = PAPTEST_CrudePrev,
    PHLTH_CrudePrev = PHLTH_CrudePrev,
    SLEEP_CrudePrev = SLEEP_CrudePrev,
    STROKE_CrudePrev = STROKE_CrudePrev,
    TEETHLOST_CrudePrev = TEETHLOST_CrudePrev
  )
```

### 2.4 Load spatial features

```{r, loading map files}
# load VB boundary census tracts
vb_boundary <-
  st_read("https://gismaps.vbgov.com/arcgis/rest/services/Basemaps/Administrative_Boundaries/MapServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json") %>%
  st_transform('EPSG:6595')

# load VB tracts
vb_tracts <- 
  st_read("https://opendata.arcgis.com/datasets/82ada480c5344220b2788154955ce5f0_2.geojson") %>%
  subset(OBJECTID!= 22) %>%
  st_transform('EPSG:6595')

#load ems stations hospitals and fire stations
ems_stations <- st_read('201127_ems_geocoded.shp') %>%
  st_transform(st_crs(vb_fishnet)) %>%
  mutate(OBJECTID = osm_id) %>%
  mutate(Legend = "ems_stations")

hospitals <- st_read('Hospitals__Virginia_.shp') %>% 
  dplyr::filter(City == 'Virginia Beach') %>%
  st_transform(st_crs(vb_fishnet)) %>%
  mutate(Legend = "hospitals")

fire_stations <- st_read('Fire_Stations.shp') %>% 
  dplyr::filter(CITY == 'Virginia Beach') %>%
  st_transform(st_crs(vb_fishnet)) %>%
  mutate(Legend = "fire_stations")
```

```{r}
# create voronoi (still needs work)
bbox_polygon <- function(x) {
  bb <- sf::st_bbox(x)
  
  p <- matrix(
    c(bb["xmin"], bb["ymin"], 
      bb["xmin"], bb["ymax"],
      bb["xmax"], bb["ymax"], 
      bb["xmax"], bb["ymin"], 
      bb["xmin"], bb["ymin"]),
    ncol = 2, byrow = T
  )
  
  sf::st_polygon(list(p))
}

box <- st_sfc(bbox_polygon(vb_boundary))

v <- st_voronoi(st_union(ems_stations), box)
plot(v, col = 0)

voronoi_polygon <- (st_intersection(st_cast(v), st_union(vb_boundary)))

plot(voronoi_polygon)

#convert sfc to sf and assign ID
vb_voronoi = voronoi_polygon %>%
  st_sf %>%
  st_cast %>%
  dplyr::mutate(Voronoi_ID = row_number())

#Assign Voronoi ID for each EMS call
main_ems.sf <- st_join(main_ems.sf, vb_voronoi, left = TRUE)
```


### 2.3 load weather data
```{r, load weather data}
# Load weather data and create features
vb_weather <- 
  riem_measures(station = "NTU", date_start = "2017-01-01", date_end = "2018-03-01") %>%
  dplyr::select(valid, tmpf, p01i, sknt, relh)%>%
  replace(is.na(.), 0) %>%
  mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
  mutate(week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>%
  group_by(interval60) %>%
  summarize(Temperature = max(tmpf),
            Precipitation = sum(p01i),
            Wind_Speed = max(sknt),
            Humidity = max(relh)) %>%
  mutate(Temperature = ifelse(Temperature == 0, 42, Temperature)) %>%
  mutate(SnowPresent = ifelse(Precipitation > 0.0 & Temperature < 32.0, "Snow", "NoSnow"),
         HeavyRain = ifelse(Precipitation > 0.5, "HeavyRain", "NoHeavyRain"),
         StrongWind = ifelse(Wind_Speed > 30, "StrongWind","NoStrongWind"))

# join weather data to the main dataframe
main_ems.sf <-
  left_join(main_ems.sf, vb_weather, by="interval60")

##GET RID OF NAs
main_ems.sf <- main_ems.sf %>%
  drop_na()
```

# 3. Exploratory Analysis

### EMS calls
- scatter plots of the calls y axis(response time))


```{r}
# create fishnet
vb_fishnet <- 
  st_make_grid(vb_boundary,
               cellsize = 1000, 
               square = FALSE) %>%
  .[vb_boundary] %>% 
  st_sf() %>%
  mutate(uniqueID = rownames(.))

# create EMS calls on fishnet
vb_ems_fishnet <- 
  dplyr::select(main_ems.sf) %>%    
  mutate(countEMS = 1,
  ) %>%    
  aggregate(., vb_fishnet, sum) %>%    
  mutate(countEMS = replace_na(countEMS, 0),
         uniqueID = rownames(.),
         cvID = sample(round(nrow(vb_fishnet) / 24),          
                       size=nrow(vb_fishnet), replace = TRUE))

# map fishnet of EMS calls in VB
ggplot() +
  geom_sf(data = vb_ems_fishnet, aes(fill = countEMS), color = NA) +
  scale_fill_viridis() +
  labs(title = "Fishnet of EMS calls in Virginia Beach")
```

```{r}
ggplot(main_ems.sf, aes(x=CallVolume, y=ResponseTime)) +
  geom_point(size = .75, colour = "darkblue") +
  labs(title= "Average Response Time by Call Volume") +
  plotTheme
```
- fishnet maps of the calls distribution and responsetime

### 3.2 Response Time and Call Volume

```{r}
# plot call volume
ggplot(main_ems.sf, aes(x=CallVolume, y=ResponseTime)) +
  geom_point(size = .75, colour = "darkblue") +
  labs(title= "Average Response Time by Call Volume") +
  plotTheme
```


# 4. Model Building

# 5. Results

# 6. Conclusion
